---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidynlm

<!-- badges: start -->
<!-- badges: end -->

tidynlm provides easy access to the National Library of Medicine API from within R. The library currently allows the user to create a dataframe of documents matching a particular search query. The documents themselves can be accessed with the urls in the `url` column of the resulting dataframe.

## Installation

You can install the released version of tidynlm from [GitHub](https://github.com/Reed-Statistics/book_trackR) with:

``` r
# install.packages("devtools")
devtools::install_github("https://github.com/Reed-Statistics/book_trackR")
```
## Cholera Demo

This is a basic example which shows you how to solve a common problem:

```{r example}
library(tidynlm)
library(tidyverse)
cholera <- searchnlm(term = "cholera", retmax = 10000, output = "wide", collapse_to_first = TRUE)
cholera %>%
  mutate(dc.date = as.numeric(dc.date)) %>%
  group_by(dc.date) %>%
  count() %>%
  ungroup() %>%
  ggplot(aes(x = dc.date, y = n)) +
  geom_point() +
  xlim(1770, 2020) +
  theme_classic() +
  labs(x = "Year",
       y = "Number of Documents",
       title = "Number of Documents Involving Cholera by Year",
       subtitle = "National Library of Medicine")
```
